---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "YOUR NAME HERE"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
inspection <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- inspection |> 
    summarize(mean_compliance = mean(compliance_score), sd_compliance = sd(compliance_score), min_compliance = min(compliance_score), max_compliance = max(compliance_score))
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
inspection |> 
  ggplot() +
  geom_histogram(aes(x = compliance_score), binwidth = 5) +
  labs(
    title = "Restaurant health inspection compliance scores in Montgomery County",
    x = "Compliance score",
    y = "Number of restaurants"
  ) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1)
```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**Answer:**
Most restaurants are very compliant. The standard deviation, mean and histogram all tell us that the majority of restaurants have good compliance scores. My next reporting steps would be to take a closer look at the outliers that scored very low and ask what exactly makes up the scores and how points are taken off (what do restaurants do to get lower scores? What should restaurant-goers know about restaurants with poor compliance scores?).
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily |> 
    summarize(mean_bus = mean(bus), sd_bus = sd(bus), mean_rail = mean(rail), sd_rail = sd(rail),)
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily |>
  mutate(
    rail_zscore = (rail - mean(rail)) / sd(rail),
    bus_zscore = (bus - mean(bus)) / sd(bus)
  )
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
unusual_days <- wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(desc(abs(rail_zscore)))

cat("Number of days with unusual rail ridership:", nrow(unusual_days), 
    "out of", nrow(wmata_zscores), "total days (", 
    round(nrow(unusual_days)/nrow(wmata_zscores)*100, 2), "%)")

unusual_days |>
  head(10) |>
  select(date, weekday, rail, rail_zscore)
```
Only six of the 425 days had unusual rail ridership and all six of those days were unusually lower ridership, rather than unusually higher.

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

In the grand scheme of things, the values of the sample are relatively close to the population. The standard deviations are higher, but the means are somewhat close to the same. This suggests that random sampling can help us get a picture of what the population might look like, but it isn't completely reliable.

```{r}
set.seed(42)
sample_91 <- wmata_daily |> sample_n(91)

sample_91 |> 
    summarize(mean_bus = mean(bus), sd_bus = sd(bus), mean_rail = mean(rail), sd_rail = sd(rail),)
```

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

I think it's interesting that all of the unusual days have unusually low ridership, rather than unusually high ridership. There were no recorded days where rail ridership was two standard deviations higher than average. You could lead a story with this, talking about how there were many unusally low days, but no unusually high ones.

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- md_car_thefts |>
  mutate(
    car_theft_rate = thefts_2023/population * 10000,
  )

```

Per-capita rate is more useful than raw counts because different counties have different populations. Somewhere with a lot of residents will likely have more car thefts than somewhere with fewer residents.

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts |> 
    summarize(median_car_theft_rate = median(car_theft_rate), total_car_thefts = sum(thefts_2023),)
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

These calculations reveal that there is a very skewed distribution of car thefts in Maryland. The top highest per-capita counties are significant higher than the rest of the counties. A lede might look like this:
Some Maryland counties have significantly more car thefts per-capita than others. Here's why:
---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

42 is a pretty high standard deviation for a mean that small. This tells us that there are some pools with far lower numbers and probably some with far higher. I would prioritize the standard deviation, because the mean might be misleading when looking at the attendence numbers for individual pools.


### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I would calculate the z-scores for each of the pools, which could tell us how statistically unusual each one is compared to the mean. This is more useful than just raw attendance because it helps us look quickly at which ones stand out when comparing to the overall mean, rather than just looking at the numbers, which might not mean much to someone unfamiliar with what a good number of attendees is for these pools.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

This is peculiar because summer is not that much longer than 60 days, so if they were going sampling randomly for more than half of the summer, it might make more sense to just track numbers across the whole summer and measure the population of the data instead. I would look into how they calculated their sample and ensure that it was truly random and I would ask them from what dates are they pulling those random 60 days from? Is it from May to September, or June to August? And if each pool tracked with a different random sample, then it might be difficult to truly compare against one another. The pools that ended up with days around July 4 in their sample might end up with much better numbers than those that didn't.
---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
