---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "OLIVER MACK"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

When comparing to Pennsylvania, it's interesting that none of the schools here have 100% FARMS. 


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

There is a clear downward trend here. The data is far more clustered compared to the Pennsylvania one, where it was kind of just in a big blob, and the line of best-fit just went across the page only slightly increased. There are a number of different explanations, one could be that the school systems could operate very differently.



### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

P-value: < 0.00000000000000022
Adjusted R-squared: 0.6734
Coefficient for FARMS pct. -0.59511

Compared to the Pennsylvania model, this is absolutely worth using! The data is statistically significant, more than half of it is explained by the independent variable and there is a clear negative correlation. Everything here tells us that there is a relationship between poverty and math scores in Montgomery County.  


### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

ANSWER HERE
Out of the ten here, Stedwick (20886) has the lowest residual (-53.92) and Caderock Springs Elementary (20817) has the highest positive residual (65.17).

Caderock's zip code has nearly three times the per capita income (127,135) of Stedwick's (45,834), so that could be a big reason for the discrepancy here. This is a huge explanation for why Caderock is significantly over-performing and why Stedwick is under-performing. As a reporter, I'd ask the schools about their socioeconomic makeups and how their districting works within the zip code as well.


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

Maryland's data did support the claim, but Pennsylvania's did not. Pennsylvania had almost no trend, whereas with Maryland, there was a clear correlation between the two variables. I would ask the school board member where the data is coming from and what his evidence is for this claim. I'd also ask more about why he thinks this is the case and what some of the influencing factors might be. I'd want to visit the school with such a low percentage to see what's going on there, what kinds of families are in the neighborhood, and what the environment of the school is.

ANSWER HERE
