---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE

Eight employees recorded zero dollars, this could skew the mean. Many of the highest earners are also detailees. There are several main salaries (every 10 or 20 thousand dollars), with some employees reviving an incremental amount in between each pay grade.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
summary(wh_salaries$SALARY)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
min(wh_salaries$SALARY)
max(wh_salaries$SALARY)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE
The median is less than the mean which means that the data is positively distributed, so most staffers likely have lower salaries (comparatively) and there are a few outliers who make far more. The minimum and maximum tells us this same story. The minimum is zero, because several staff members had no reported salary, and the maximum is 225700, which is far larger than the median and mean.

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE
I would look at who in the White House earns the most and look more into the detailees specifically. I would try and write about the top earners and figure out why they'd be earning so much more. I would start by looking into the names of each of the top earners (and some of the earners throughout the distribution) and then I would try to figure out how long each person on this list has worked at the White House.

