---
title: "HW9: Car Theft Confidence Intervals & Margin of Error"
name: OLIVER MACK
---

Your editor wants to know what a "typical" month of car thefts looks like in Prince George's County. Not just the average — she wants to know how confident you can be in that number. Using monthly figures, confidence intervals and margin of error, you'll come up with an answer and decide whether the data supports a clear story. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# LOAD THE TIDYVERSE
```

### Task 2: Load the Prince George's county crime data

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. What's your initial reaction to the change in car thefts? Is there a trend? If your editor asked you "how many car thefts happen in a typical month?", what would you say based on this chart alone?

There is a strong downward trend over the years. If my editor asked me that for today, I would say between 200 and 400. If we're looking at the average across the whole chart, I would probably say somewhere between 350 and 450.

```{r}

car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2026)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```

### Task 4: Calculate the mean, standard error and margin of error

We've got a small number of months, so we need to quantify the uncertainty around our average. We'll calculate the standard error (which measures how much the sample mean could vary), and the margin of error (which gives us the plus/minus range). Replace the REPLACE_ME values based on what we did in class.

```{r}

theft_summary <- car_thefts |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts
  )

# Display the results
theft_summary

```

**Answer this question:** Explain in your own words what the margin of error tells us here. If you were writing a story, how would you use this number? For example, how would you complete this sentence: "Prince George's County averages about ___ car thefts per month, plus or minus ___."

I would say something along the lines of: Prince George's County averages about 376 car thefts per month, plus or minus about 34.

### Task 5: Build the confidence interval

Now let's turn the margin of error into a full confidence interval — the range where we're 95% confident the true average falls. Replace the REPLACE_ME values.

```{r}

theft_summary <- theft_summary |>
  mutate(
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

theft_summary |>
  select(mean_thefts, margin_error, ci_lower, ci_upper)

```

**Answer this question:** If a county official claimed "car thefts are down to about 300 per month," does your confidence interval support that claim? What about a claim of 350 per month? Explain how the confidence interval helps you evaluate each claim.

The confidence interval does not support the claim that they're down to about 300 per month, because that number is completely outside of the confidence interval. About 350 car thefts would be acceptable, but it would be more accurate to say about 375.  The confiende interval helps us figure out if the cliam is a reasonable number based on the margin of error. A number like 300 is entirely outside of the margin of error, so it wouldn't be accurate.

### Task 6: Visualize the data with the confidence interval

Let's make a chart showing the data with the mean and the 95% confidence interval. Does this change anything about your answer from Task 3?

It doesn't particularly change my answer. There is still a downward trend, as the number of car thefts has been below/around the mean since late 2024. I said if I was looking at the whole chart, I would place the average between 350 and 450, and this is nearly true, just a little bit wider than the actual mean/margin of error.

```{r}
# Make a chart with the mean and CI
car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper),
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month),
           y = theft_summary$mean_thefts + 10,
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)),
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 7: The margin of error and sample size

In class we learned that MoE = 1/√n × 100 as a quick approximation for polls. That same principle applies here: more data means a smaller margin of error. Let's see what would happen if we had 40 months of data instead of the number we have, assuming the same mean and standard deviation. Replace the REPLACE_ME value with 40.

```{r}

hypothetical_summary <- tibble(
  mean_thefts = theft_summary$mean_thefts,
  sd_thefts = theft_summary$sd_thefts,
  months = 40,
  se_thefts = sd_thefts / sqrt(months),
  margin_error = qt(0.975, df = months - 1) * se_thefts,
  ci_lower = mean_thefts - margin_error,
  ci_upper = mean_thefts + margin_error
)

# Compare the two
comparison <- tibble(
  scenario = c("Actual data", "40 months"),
  months = c(theft_summary$months, hypothetical_summary$months),
  margin_error = c(theft_summary$margin_error, hypothetical_summary$margin_error),
  ci_lower = c(theft_summary$ci_lower, hypothetical_summary$ci_lower),
  ci_upper = c(theft_summary$ci_upper, hypothetical_summary$ci_upper)
)

comparison

```

**Answer this question:** How did the margin of error and confidence interval change? What does this tell you about the tradeoff between having more data and being able to make precise claims? If you were covering crime trends, how many months of data would you want before writing a definitive story?

The margin of error went down, and the confidence interval lower and upper boundaries got closer together. This shows that the more months we have, the more precise our claims become. I suppose it depends on the type of story, but it would be nice to have at least three years (36 months) of data to show a trend, but the more the better!

### Task 8: Reporting with uncertainty

Based on everything you've calculated, answer the following:

1. Write a headline and a one-paragraph lede about car thefts in Prince George's County. Your text should properly reflect the uncertainty in the data — don't overstate or understate what the numbers show.

2. Your editor says "Just tell me if car thefts are going up or down." Using only the confidence interval (no significance testing yet), what's the most accurate answer you can give? What would you need to do next to answer that question more definitively?

PUT ANSWERS HERE

1: HL: Prince George's County car thefts decline in recent months
STORY: Prince George's County averaged about 375 car thefts over the past three years, give or take about 34. In recent months, the county has seen a decline in car thefts. Last month saw the fewest car thefts since mid 2023, at just over 200. 

2: I would tell the editor that car thefts are going down. This is because for more than a year, they've been around or below the mean and many of the most recent points are even below the lower confidence interval boundary. To answer the question more definitively, I would like to see a bit more data and calculate exactly how far away each data point is from the mean. 

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
