---
title: "final_project"
output: html_document
---
Below I've loaded two datasets. The first is crime_data_states, which is crime data on each state from 1960 through 2019. This data is from the CORGIS Dataset Project by Ryan Whitcomb, Joung Min Choi and Bo Guan. It was created in 2021 and it pulls data directly from the FBI's crime statistics page. In this dataset, I think the most important columns are going to be the overall property rate as well as the overall violent rate. I plan to use these columns to directly compare the rates to their respective state unemployment rates. I plan to split this up by year as well, so that we can get a clearer picture of trends over time (if any exist!) This data covers every state from 1960 to 2019. The second is a dataset is compiled data from the U.S. Bureau of Labor Statistics from 1976-2024.

```{r load data}
library(tidyverse)
library(dplyr)

crime_data_states <- read_csv("https://raw.githubusercontent.com/oliver-mack/jour405_spring26/refs/heads/main/state_crime%20(1).csv")
unemployment_states <- read_csv("https://raw.githubusercontent.com/oliver-mack/jour405_spring26/refs/heads/main/Untitled%20spreadsheet%20-%20staadata.csv")

```

Below is the code that Prof. Willis provided that helps clean up the unemployment_states data.

```{r}
#this is to clean up the data!
unemployment_states <- read_csv("https://raw.githubusercontent.com/oliver-mack/jour405_spring26/refs/heads/main/Untitled%20spreadsheet%20-%20staadata.csv") |>
  filter(!is.na(`FIPS Code`), 
         nchar(str_trim(as.character(`FIPS Code`))) == 2) |>
  mutate(across(where(is.character), str_trim)) 
```

Below are several codes I used to clean up the data a bit and combine the two data sets.

```{r}
#this is to rename the year variable to match the other dataset.
unemployment_states <- unemployment_states %>%
  rename(Year = year)
```
```{r}
#This is to rename the state/area variable
unemployment_states <- unemployment_states %>% 
rename(State = 'state/area')
```
```{r}
#this is to combine the two datasets
combined <- crime_data_states |> inner_join(unemployment_states, join_by(Year, State))
```

First, I'll just do summaries of each column, across the entire dataset (every year).
```{r}
summary(combined$`Unemployment-Rate`)
```
```{r}
summary(combined$`Data.Rates.Violent.All`)
```

We'll look briefly at the top few, just out of curiosity.
```{r}
combined |> arrange(desc(`Unemployment-Rate`)) |> head(10)
```

It's interesting that West Virginia and Michigan alone take up seven of the top ten spots, with Alabama taking up another two. We'll do the same for violent crime rates.

```{r}
combined |> arrange(desc(`Data.Rates.Violent.All`)) |> head(10)
```
Interesting here that Washington, D.C., has all top ten spots, but this makes sense, as D.C. is just entirely city and lacks much suburban/urban areas like every other state has.

Then, I made an overall visualization of the data across all years.

```{r}
combined |>
  ggplot(aes(x = `Unemployment-Rate`, y = `Data.Rates.Violent.All`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Unemployment Rate", y = "Violent Crime Rate",
       title = "Unemployment Rate vs. Violent Crime Rate")
```
Since there are 2,244 entries, this upward trend is really interesting to me. Though the data is kind of in a large clump, we can tell that there is the possibility of some type of relationship existing between the two, just due to the fact that there is a slight positive correlation and there are so many different data points. Next, I want to try and make the graph a bit more clear by elimintaing the outliars, so I've removed any data points that have crime rates above 1500.

```{r}
combined |>
  filter(`Data.Rates.Violent.All` <= 1500) |>
  ggplot(aes(x = `Unemployment-Rate`, y = `Data.Rates.Violent.All`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Unemployment Rate", y = "Violent Crime Rate",
       title = "Unemployment Rate vs. Violent Crime Rate")
```

There is a relationship here, though the data points are also just put into a very large clump. Next, I'll check the model.

```{r}
model <- lm(`Data.Rates.Violent.All` ~ `Unemployment-Rate`, data = combined)
summary(model)
```
The P-value is very low and we do have a pretty high coefficient. Our R-squared is pretty low though. So there is a relationship here, but it doesn't seem to be a very strong or predictive. It seems like there is some sort of postiive relationship, but it's likely determined by other factors. I want to check a few specific years as well.

```{r}
combined |>
  filter(Year == 2019) |>
  ggplot(aes(x = `Unemployment-Rate`, y = `Data.Rates.Violent.All`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Unemployment Rate", y = "Violent Crime Rate",
       title = "Unemployment Rate vs. Violent Crime Rate")
```
Here, it looks like we might have a stronger relationship, although, the datapoints are still rather clumped, and it seems that the three outliers, each above 750, could be controlling that line of best fit again. Let's look at a model from 2015.

```{r}
model <- lm(`Data.Rates.Violent.All` ~ `Unemployment-Rate`, data = combined[combined$Year == 2015, ])
summary(model)
```

For the model, I found how to do a specific year filter online and it looks here that we have a bit more predictability (28%), but it's still not a very strong relationship. The data also looks quite blob-ish, which could indicate a weak relationship even though there is a positive line. Let's try just one state next!

```{r}
combined |>
  filter(State == 'Maryland') |>
  ggplot(aes(x = `Unemployment-Rate`, y = `Data.Rates.Violent.All`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Unemployment Rate", y = "Violent Crime Rate",
       title = "Unemployment Rate vs. Violent Crime Rate")
```

```{r}
model <- lm(`Data.Rates.Violent.All` ~ `Unemployment-Rate`, data = combined[combined$State == 'Maryland', ])
summary(model)
```
Somehow, the adjusted R-squared is in the negatives, and I think that the ggplot shows us that there is virtually no relationship here. The p-value is far above the statistically significant level and the data is all over the place on the chart. 

Next, I want to look into the delay a bit more, so here is a code that Prof. Willis sent to help with that!

```{r}
# Create a lagged unemployment variable
maryland_lagged <- combined |>
  filter(State == "Maryland") |>
  arrange(Year) |>
  mutate(unemployment_lag1 = lag(`Unemployment-Rate`, n = 1))

# Now correlate lagged unemployment with crime rate
cor.test(maryland_lagged$unemployment_lag1, 
         maryland_lagged$`Data.Rates.Violent.All`, 
         use = "complete.obs")

# Or run a regression with the lagged predictor
model_lag <- lm(`Data.Rates.Violent.All` ~ unemployment_lag1, 
                data = maryland_lagged)
summary(model_lag)
```

Even with the lag filter, we see very little statistical significance between the two variables by looking at the p-value and adjusted r-squared, which is still below zero.

FINAL SECTION:

I wouldn't pitch this story until doing more reporting on it. If I had to write a story just on the data I found here, I would go through each state and look for states with the largest changes overtime, or perhaps, the largest correlations. It could be interesting to write about how in Maryland, the amount of unemployment doesn't seem to affect the violent crime rate much at all over as one or the other changes over time. Another story here could be 

My main findings are that though there appears to be a positive relationship between unemployment and violent crime, the two variables are largely unconnected. While there is a relationship between the two of them, and we can see a positive relationship, our models tell us that unemployment is not much of an explaining factor in the violent crime rate. My initial prediction is that the two would have a relationship, but not be the sole determining factor, which is true, but to more of an extreme extent.

Based on the adjusted R-squared we found, the  between In terms of a reporting plan, I believe that it would be best to continue testing other variables to see what might would have stronger relationships with crime rates. I would start doing wide tests of other variables, such as education levels, median income, poverty rate, etc. 

